---
title: "DATA607 Final Project Code"
author: "Rick Powell"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
bibliography: references.bib
nocite: '@*'
---

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# This extracts worldwide COVID19 data by state
# install.packages("COVID19") <-- package already installed
library("COVID19")
library(tidyverse)
library(sf)
library(dplyr)
library(pheatmap)
library(RColorBrewer)
```


# Intro

Originally this was my Storyboard, however, I've made a bunch of changes so I wanted to keep the graphs and information here, but some of the graphs and/or text either doesn't apply, or is outdated.

```{r, echo=TRUE}
df_state <- covid19(level = 2, verbose = FALSE)
df_country <- covid19(level = 1, verbose = FALSE)
df_us_states <- subset(df_state, administrative_area_level_1 == "United States")
```

# Cleaning to get the Columns we want to look at for the US States.

```{r, echo=TRUE, warning=FALSE}
df_us_states <- df_us_states %>%
  select(date, confirmed, deaths, recovered, tests, vaccines, 
         school_closing, workplace_closing, stay_home_restrictions, 
         facial_coverings, administrative_area_level_1, 
         administrative_area_level_2, population)

# Take the absolute value of these columns and replace original values
columns_to_fix <- c("school_closing", "workplace_closing", "stay_home_restrictions", "facial_coverings")

df_us_states <- df_us_states %>%
  mutate(across(all_of(columns_to_fix), abs))
```

```{r, echo=TRUE, warning=FALSE}
# Remove these territories from the data set
df_us_states <- df_us_states %>%
  filter(!administrative_area_level_2 %in% c("Northern Mariana Islands", 
                                             "Virgin Islands", 
                                             "Guam", 
                                             "American Samoa", 
                                             "Puerto Rico"))
```


# Country Level

```{r, echo=TRUE, warning=FALSE}
# Filter data for the United States
df_filtered <- df_country[df_country$administrative_area_level_1 == "United States",]

# Create time series plot
ggplot(df_filtered, aes(x = date, y = confirmed)) +
  geom_line(color = "blue") + 
  labs(title = "Confirmed Cases Over Time in the United States", 
       x = "Date", 
       y = "Confirmed Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_labels = "%Y",
               date_breaks = "1 year",
               limits = as.Date(c("2020-01-01", "2023-06-30")))
```

Starting with Cases over Time for US. The focus of this project will be on the US, so I want to set some baseline of how the US compares overall. I probably won't use this first graph, but it's nice to have.


```{r, echo=TRUE, warning=FALSE}
# Create time series plot
ggplot(df_filtered, aes(x = date, y = deaths)) +
  geom_line(color = "blue") + 
  labs(title = "Covid Deaths Over Time in the United States", 
       x = "Date", 
       y = "Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_labels = "%Y",
               date_breaks = "1 year",
               limits = as.Date(c("2020-01-01", "2023-06-30")))
```

```{r, echo=TRUE, warning=FALSE, fig.width=10, fig.height=4}
# Filter the data for the United States
df_filtered <- df_country[df_country$administrative_area_level_1 == "United States",]

# Reshape data into long format
df_long <- df_filtered %>%
  select(date, confirmed, deaths, vaccines) %>%
  pivot_longer(
    cols = c(confirmed, deaths, vaccines),
    names_to = "metric",
    values_to = "count"
  )

# Create faceted plot
ggplot(df_long, aes(x = date, y = count)) +
  geom_line(aes(color = metric), size = 1) +
  labs(
    title = "COVID-19 Metrics Over Time in the United States",
    x = "Date",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none" 
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(
    date_labels = "%Y",
    date_breaks = "1 year",
    limits = as.Date(c("2020-01-01", "2023-06-30"))
  ) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  scale_color_manual(
    values = c("confirmed" = "blue", "deaths" = "red", "vaccines" = "green")
  )
```


# State Level

```{r, echo=TRUE, warning=FALSE}
ggplot(df_us_states, aes(x = date, y = confirmed, color = administrative_area_level_2)) +
  geom_line() +  
  labs(title = "Confirmed Covid Cases Over Time by US State", 
       x = "Date", 
       y = "Confirmed Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
       legend.position = "none" ) +
  scale_x_date(date_labels = "%Y",
               date_breaks = "1 year",
               limits = as.Date(c("2020-01-01", "2023-06-30")))
```

```{r, echo=TRUE, warning=FALSE}
ggplot(df_us_states, aes(x = date, y = (confirmed / population), color = administrative_area_level_2)) +
  geom_line() +  
  labs(title = "Confirmed Covid Cases by Population Over Time by US State", 
       x = "Date", 
       y = "Confirmed Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
       legend.position = "none" ) +
  scale_x_date(date_labels = "%Y",
               date_breaks = "1 year",
               limits = as.Date(c("2020-01-01", "2023-06-30")))
```

Here's that same graph showing cases over population. Fewer standouts, but very cluttered.

```{r, echo=TRUE, warning=FALSE}
# Calculate the US average for confirmed cases per population by date
us_average <- aggregate((confirmed / population) ~ date, data = df_us_states, mean, na.rm = TRUE)
colnames(us_average) <- c("date", "confirmed_per_population")

# Update state group for coloring (highlight selected states)
highlight_states <- c("North Carolina")
df_us_states$state_group <- ifelse(df_us_states$administrative_area_level_2 %in% highlight_states, 
                                    df_us_states$administrative_area_level_2, "Other")

# Create the plot
ggplot() +
  # Add state-level lines
  geom_line(data = df_us_states, 
            aes(x = date, y = (confirmed / population), 
                color = state_group, group = administrative_area_level_2), 
            size = 1, alpha = 0.7) +
  # Labels and theme
  labs(title = "Confirmed Covid Cases by Population Over Time by US State", 
       x = "Date", 
       y = "Confirmed Cases per Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(date_labels = "%Y",
               date_breaks = "1 year",
               limits = as.Date(c("2020-01-01", "2023-06-30"))) +
  # Custom colors for states and US average
  scale_color_manual(values = c("North Carolina" = "blue", 
                                 "Other" = "gray"),
                     breaks = c("North Carolina"),
                     labels = c("North Carolina")) +
  guides(color = guide_legend(title = "State"))
```
Originally this graph showed the US Average with a couple of the high and low data, but I re-used it to see North Carolina, which appeared as a stand alone cluster in an earlier try. I'm not using this.

```{r, echo=TRUE, warning=FALSE}
df_us_states$facial_coverings_group <- factor(df_us_states$facial_coverings,
                                              levels = 0:4,
                                              labels = c("No Policy", 
                                                         "Recommended", 
                                                         "Required in Some Spaces", 
                                                         "Required in All Shared Spaces", 
                                                         "Required Outside Home"))

july_and_jan_data <- df_us_states %>%
  filter(date %in% as.Date(c("2020-07-01", "2021-01-01", "2021-07-01", 
                            "2022-01-01", "2022-07-01"))) %>%
  select(administrative_area_level_2, date, facial_coverings_group)

# Load map data for U.S. states
us_states <- map_data("state") 

# Convert state names to lowercase
july_and_jan_data$administrative_area_level_2 <- tolower(july_and_jan_data$administrative_area_level_2)

# Merge policy data with map data
map_data_combined <- us_states %>%
  left_join(july_and_jan_data, by = c("region" = "administrative_area_level_2"))

# Create the faceted map plot
ggplot(map_data_combined, aes(x = long, y = lat, group = group, fill = facial_coverings_group)) +
  geom_polygon(color = "black") +
  coord_fixed(1.3) + 
  labs(title = "Facial Coverings Policy During Covid",
       fill = "Facial Coverings Policy",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  scale_fill_manual(values = c("No Policy" = "darkgreen",
                                "Recommended" = "skyblue",
                                "Required in Some Spaces" = "blue",
                                "Required in All Shared Spaces" = "orange",
                                "Required Outside Home" = "red")) +
  facet_wrap(~date, ncol = 2)
```


```{r, echo=TRUE, warning=FALSE}
# Filter data for July 1st and January 1st again, but for school closing policies
school_closing_data <- df_us_states %>%
  filter(date %in% as.Date(c("2020-07-01", "2021-01-01", "2021-07-01", 
                             "2022-01-01", "2022-07-01"))) %>%
  select(administrative_area_level_2, date, school_closing)

# Convert state names to lowercase
school_closing_data$administrative_area_level_2 <- tolower(school_closing_data$administrative_area_level_2)

# Add a grouping column for school closing
school_closing_data$school_closing_group <- factor(school_closing_data$school_closing,
                                                   levels = 0:3,
                                                   labels = c("No Measures", 
                                                              "Recommend Closing or Alterations", 
                                                              "Require Closing (Some Levels)", 
                                                              "Require Closing (All Levels)"))

# Merge with map data
map_data_school <- us_states %>%
  left_join(school_closing_data, by = c("region" = "administrative_area_level_2"))

# Create faceted map plot
ggplot(map_data_school, aes(x = long, y = lat, group = group, fill = school_closing_group)) +
  geom_polygon(color = "black") +
  coord_fixed(1.3) +
  labs(title = "School Closing Policies During Covid",
       fill = "School Closing Policy",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  scale_fill_manual(values = c("No Measures" = "darkgreen",
                                "Recommend Closing or Alterations" = "skyblue",
                                "Require Closing (Some Levels)" = "orange",
                                "Require Closing (All Levels)" = "red")) +
  facet_wrap(~date, ncol = 2)

```


```{r, echo=TRUE, warning=FALSE}
# Filter data for July 1st and January 1st again, but for workplace closing policies
workplace_closing_data <- df_us_states %>%
  filter(date %in% as.Date(c("2020-07-01", "2021-01-01", "2021-07-01", 
                             "2022-01-01", "2022-07-01"))) %>%
  select(administrative_area_level_2, date, workplace_closing)

# Convert state names to lowercase
workplace_closing_data$administrative_area_level_2 <- tolower(workplace_closing_data$administrative_area_level_2)

# Add a grouping column for workplace closing
workplace_closing_data$workplace_closing_group <- factor(workplace_closing_data$workplace_closing,
                                                         levels = 0:3,
                                                         labels = c("No Measures", 
                                                                    "Recommend Closing or Alterations", 
                                                                    "Require Closing (Some Sectors)", 
                                                                    "Require Closing (All-But-Essential)"))

# Merge with map data
map_data_workplace <- us_states %>%
  left_join(workplace_closing_data, by = c("region" = "administrative_area_level_2"))

# Create faceted map plot
ggplot(map_data_workplace, aes(x = long, y = lat, group = group, fill = workplace_closing_group)) +
  geom_polygon(color = "black") +
  coord_fixed(1.3) +
  labs(title = "Workplace Closing Policies During Covid",
       fill = "Workplace Closing Policy",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  scale_fill_manual(values = c("No Measures" = "darkgreen",
                                "Recommend Closing or Alterations" = "skyblue",
                                "Require Closing (Some Sectors)" = "orange",
                                "Require Closing (All-But-Essential)" = "red")) +
  facet_wrap(~date, ncol = 2)
```
```{r, echo=TRUE, warning=FALSE, fig.width=8, fig.height=8}
# Filter the data for July 1st, 2020
snapshot_data <- df_us_states %>%
  filter(date == as.Date("2020-07-01")) %>%
  select(administrative_area_level_2, confirmed, deaths, school_closing, workplace_closing, stay_home_restrictions, facial_coverings)

# Make sure categorical variables are numeric
snapshot_data <- snapshot_data %>%
  mutate(
    school_closing = as.numeric(as.factor(school_closing)),
    workplace_closing = as.numeric(as.factor(workplace_closing)),
    stay_home_restrictions = as.numeric(as.factor(stay_home_restrictions)),
    facial_coverings = as.numeric(as.factor(facial_coverings))
  )

# Set the states as row names
heatmap_data <- snapshot_data %>%
  column_to_rownames(var = "administrative_area_level_2")


# Create the heatmap
pheatmap(
  heatmap_data,
  scale = "column",
  cluster_rows = T, cluster_cols = T,
  angle_col = 45,
  show_colnames = T, show_rownames = F,
  main = "Heatmap of COVID-19 Metrics on July 1st, 2020",
  col = brewer.pal(10, 'RdYlBu'),
  legend = FALSE
)
```
I really wanted to use a heatmap, especially because it was what we did our presentation on, but it just didn't work out that way. I still want to do some sort of clustering

```{r, echo=TRUE, warning=FALSE, fig.width=8}
# Filter the data for July 1st, 2020 and select only the 4 relevant columns
snapshot_data <- df_us_states %>%
  filter(date == as.Date("2020-07-01")) %>%
  select(administrative_area_level_2, school_closing, workplace_closing, stay_home_restrictions, facial_coverings)

# Make sure categorical variables are numeric
snapshot_data <- snapshot_data %>%
  mutate(
    school_closing = as.numeric(as.factor(school_closing)),
    workplace_closing = as.numeric(as.factor(workplace_closing)),
    stay_home_restrictions = as.numeric(as.factor(stay_home_restrictions)),
    facial_coverings = as.numeric(as.factor(facial_coverings))
  )

# Remove the "administrative_area_level_2" column and set states as row names
heatmap_data <- snapshot_data %>%
  select(-administrative_area_level_2) %>%
  as.data.frame()

# Set row names to the states (administrative_area_level_2)
rownames(heatmap_data) <- snapshot_data$administrative_area_level_2

# Perform hclust on the rows based on the 4 policies
hclust_result <- hclust(dist(heatmap_data))

# Plot the dendrogram with the state names
plot(hclust_result, 
     main = "Dendrogram of US States (Policies)", 
     xlab = "", 
     sub = "", 
     cex = 0.9, 
     labels = rownames(heatmap_data))
```
This is a basic clustering to test my idea. Now I scale it so we have actual better clusters.


```{r, echo=TRUE, warning=FALSE, fig.width=8}
# Filter the data for the specified dates and select only the relevant columns
snapshot_data <- df_us_states %>%
  filter(date %in% as.Date(c("2020-07-01", "2021-01-01", "2021-07-01", "2022-01-01", "2022-07-01"))) %>%
  select(administrative_area_level_2, date, school_closing, workplace_closing, stay_home_restrictions, facial_coverings)

# Make sure categorical variables are numeric
snapshot_data <- snapshot_data %>%
  mutate(
    school_closing = as.numeric(as.factor(school_closing)),
    workplace_closing = as.numeric(as.factor(workplace_closing)),
    stay_home_restrictions = as.numeric(as.factor(stay_home_restrictions)),
    facial_coverings = as.numeric(as.factor(facial_coverings))
  )

# Reshape the data so that each state only appears once, with columns for each date's values
snapshot_data_wide <- snapshot_data %>%
  pivot_wider(
    names_from = date,
    values_from = c(school_closing, workplace_closing, stay_home_restrictions, facial_coverings),
    names_glue = "{.value}_{format(date, '%Y-%m-%d')}"
  )

# Remove the "administrative_area_level_2" column for clustering
heatmap_data <- snapshot_data_wide %>%
  select(-administrative_area_level_2)

# Perform hclust on the rows based on the 4 policies across the 5 dates
hclust_result <- hclust(dist(heatmap_data))

# Plot the dendrogram with the state names
plot(hclust_result, 
     main = "Clustering of US States Based on COVID-19 Policies", 
     xlab = "", 
     sub = "", 
     cex = 0.9, 
     labels = snapshot_data_wide$administrative_area_level_2)
```


```{r, echo=TRUE, warning=FALSE}
# Calculate the US average for confirmed cases per population by date
us_average <- aggregate((confirmed / population) ~ date, data = df_us_states, mean, na.rm = TRUE)
colnames(us_average) <- c("date", "confirmed_per_population")

# Update state group for coloring 
highlight_states <- c("Alaska", "American Samoa", "Rhode Island")
df_us_states$state_group <- ifelse(df_us_states$administrative_area_level_2 %in% highlight_states, 
                                    df_us_states$administrative_area_level_2, "Other")

# Create the plot
ggplot() +
  # Add state-level lines
  geom_line(data = df_us_states, 
            aes(x = date, y = (confirmed / population), 
                color = state_group, group = administrative_area_level_2), 
            size = 1, alpha = 0.7) +
  # Add US average line
  geom_line(data = us_average, 
            aes(x = date, y = confirmed_per_population, color = "United States"), 
            size = 1) +
  # Labels and theme
  labs(title = "Confirmed Covid Cases by Population Over Time by US State", 
       x = "Date", 
       y = "Confirmed Cases per Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(date_labels = "%Y",
               date_breaks = "1 year",
               limits = as.Date(c("2020-01-01", "2023-06-30"))) +
  # Custom colors for states and US average
  scale_color_manual(values = c("Alaska" = "blue", 
                                 "American Samoa" = "coral", 
                                 "Rhode Island" = "darkgreen", 
                                 "United States" = "purple", 
                                 "Other" = "gray"),
                     breaks = c("Alaska", "American Samoa", "Rhode Island", "United States"),
                     labels = c("Alaska", "American Samoa", "Rhode Island", "US Average")) +
  guides(color = guide_legend(title = "State"))
```
Here's that graph I mentioned earlier with the US Average and a couple high confirmed/population states. Turns out I copied it down here. This was before I remove the territories, so American Samoa is still in there as it was way below every other state/territory when I originally ran this.

```{r, echo=TRUE, warning=FALSE}
# Load a state shapefile
states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))

# Create a clustering data frame
state_clusters <- data.frame(
  state = tolower(c(
    "North Carolina", 
    "Hawaii", "Louisiana", "Colorado", "California", "New York", 
    "Maryland", "District of Columbia",
    "Oklahoma", "Kansas", "New Hampshire", "North Dakota", "Iowa",
    "West Virginia", "Kentucky", "Texas", "Tennessee", "Idaho",
    "Illinois", "Minnesota", "Nevada", "Mississippi", "South Carolina",
    "Florida", "Missouri", "Arizona", "Montana", "Wyoming", "South Dakota",
    "Wisconsin", "Alabama", "Ohio", "New Jersey", "Maine", "Virginia",
    "Indiana", "Georgia", "Nebraska", "Utah", "Alaska", "Oregon", 
    "New Mexico", "Washington", "Connecticut", "Arkansas", "Delaware",
    "Massachusetts", "Vermont", "Michigan", "Rhode Island", "Pennsylvania"
  )),
  cluster = c(
    1, 
    rep(2, 7), 
    rep(3, 10), 
    rep(4, 13), 
    rep(5, 8), 
    rep(6, 12)
  )
)


# Merge state clusters with the shapefile
states$cluster <- state_clusters$cluster[match(states$ID, state_clusters$state)]

# Plot with ggplot2
ggplot(states) +
  geom_sf(aes(fill = as.factor(cluster)), color = "black", size = 0.1) +
  scale_fill_brewer(palette = "Set3", name = "Cluster") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(
    title = "Clustering of US States"
  )
```
Originally, I just wanted to see what 6 clusters would look like


```{r, echo=TRUE, warning=FALSE}
silhouette_score <- function(k) {
  clustering <- cutree(hclust(dist(heatmap_data)), k)
  silhouette_result <- silhouette(clustering, dist(heatmap_data))
  mean(silhouette_result[, 3])  # Average silhouette width
}

silhouette_values <- sapply(2:20, silhouette_score)

# Create a data frame with k values and corresponding silhouette scores
silhouette_data <- data.frame(
  k = 2:20,
  silhouette_width = silhouette_values
)

# Create the plot
ggplot(silhouette_data, aes(x = k, y = silhouette_width)) +
  geom_point(aes(color = k == 2), size = 3) +  
  scale_color_manual(values = c("black", "red")) + 
  geom_line() +  
  labs(
    x = "Number of Clusters (k)",
    y = "Average Silhouette Width",
    title = "Silhouette Method for Optimal k"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
But then it made sense to find the actual best number of clusters, which is 2.

```{r, echo=TRUE, warning=FALSE}
# Optimal k = 2
optimal_k <- 2
clustering <- cutree(hclust(dist(heatmap_data)), optimal_k)

# Compute silhouette scores
silhouette_result <- silhouette(clustering, dist(heatmap_data))

# Create a data frame with states and silhouette scores
low_silhouette <- data.frame(
  state = rownames(heatmap_data),
  silhouette_score = silhouette_result[, 3]
)

# Filter states with low silhouette scores (threshold < 0.1)
low_confidence_states <- low_silhouette %>%
  filter(silhouette_score < 0.1) %>%
  arrange(silhouette_score)

# View the result
print(low_confidence_states)
```

```{r, echo=TRUE, warning=FALSE}
# Filter the data for the specified dates and select only the relevant columns
snapshot_data <- df_us_states %>%
  filter(date %in% as.Date(c("2020-07-01", "2021-01-01", "2021-07-01", "2022-01-01", "2022-07-01"))) %>%
  select(administrative_area_level_2, date, school_closing, workplace_closing, stay_home_restrictions, facial_coverings)

# Make sure categorical variables are numeric
snapshot_data <- snapshot_data %>%
  mutate(
    school_closing = as.numeric(as.factor(school_closing)),
    workplace_closing = as.numeric(as.factor(workplace_closing)),
    stay_home_restrictions = as.numeric(as.factor(stay_home_restrictions)),
    facial_coverings = as.numeric(as.factor(facial_coverings))
  )

# Reshape the data so that each state only appears once, with columns for each date's values
snapshot_data_wide <- snapshot_data %>%
  pivot_wider(
    names_from = date,
    values_from = c(school_closing, workplace_closing, stay_home_restrictions, facial_coverings),
    names_glue = "{.value}_{format(date, '%Y-%m-%d')}"
  )

# Remove the "administrative_area_level_2" column for clustering
heatmap_data <- snapshot_data_wide %>%
  select(-administrative_area_level_2)

# Perform hclust on the rows (states) based on the 4 policies across the 5 dates
hclust_result <- hclust(dist(heatmap_data))

# Cut the dendrogram to obtain 2 clusters
clusters <- cutree(hclust_result, k = 2)

# Add the cluster labels to the original data
snapshot_data_wide$cluster <- clusters

# Plot the dendrogram with the state names
plot(hclust_result, 
     main = "Clustering of US States Based on COVID-19 Policies", 
     xlab = "", 
     sub = "", 
     cex = 0.9, 
     labels = snapshot_data_wide$administrative_area_level_2)

# Display the resulting clusters
snapshot_data_wide %>%
  select(administrative_area_level_2, cluster)

```
This helped me see what our clustering looks like with our optimal 2 clusters.

```{r, echo=TRUE, warning=FALSE}
# Adjust margins for a better view
par(mar = c(2, 4, 4, 2))

# Rescale the y-axis and plot the dendrogram
plot(hclust_result, 
     main = "Clustering of US States Based on COVID-19 Policies", 
     xlab = "", 
     sub = "", 
     cex = 0.7,                    
     labels = snapshot_data_wide$administrative_area_level_2,
     hang = -1,                    
     ylim = c(0, max(hclust_result$height) * 2), 
     yaxt = "n")                   

# Add a larger y-axis
axis(2, at = seq(0, max(hclust_result$height) * 2, by = 1), las = 1, cex.axis = 0.8)
```


```{r, echo=TRUE, warning=FALSE}
# Define the state names and their respective clusters
state_clusters <- data.frame(
  state = tolower(c(
    "Minnesota", "California", "Florida", "Wyoming", "South Dakota", "Kansas",
    "Nevada", "Virginia", "Washington", "Oregon", "Wisconsin", "New Jersey",
    "Rhode Island", "Vermont", "North Carolina", "Oklahoma", "Alabama", "Delaware",
    "Missouri", "Utah", "Mississippi", "Connecticut", "Indiana", "Georgia", "Texas",
    "Pennsylvania", "Massachusetts", "Maine", "Tennessee", "Michigan", "Idaho", "Illinois",
    "Louisiana", "New Mexico", "Arizona", "Arkansas", "Nebraska", "West Virginia", "South Carolina",
    "New York", "District of Columbia", "Kentucky", "Ohio", "Alaska", "New Hampshire", "North Dakota",
    "Iowa", "Montana", "Hawaii", "Maryland", "Colorado"
  )),
  cluster = c(
    1, 2, 1, 1, 1, 1, 1, 2, 2, 2, 1, 2, 2, 2, 1, 1, 1, 2, 1, 2, 1, 2, 2, 2, 1, 2, 2, 
    2, 1, 2, 1, 1, 2, 2, 1, 2, 2, 1, 1, 2, 2, 1, 2, 2, 1, 1, 1, 1, 2, 2, 2
  )
)

# Load the shapefile for US states
states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))

# Merge state clusters with the shapefile
states$cluster <- state_clusters$cluster[match(tolower(states$ID), state_clusters$state)]

# Plot the map
ggplot(states) +
  geom_sf(aes(fill = as.factor(cluster)), color = "black", size = 0.1) +
  scale_fill_manual(values = c("coral", "skyblue"), name = "Cluster") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(
    title = "Clustering of US States Based on COVID-19 Policies"
  )
```
And here's our final clustering!


```{r, echo=TRUE, warning=FALSE}
# Load required libraries
library(sf)
library(ggplot2)
library(dplyr)

# Load the shapefile for US states using the `maps` package
states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))

# Create a data frame with voting results
state_votes <- data.frame(
  region = tolower(c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", 
                     "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", 
                     "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", 
                     "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
                     "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", 
                     "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
                     "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", 
                     "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", 
                     "Washington", "West Virginia", "Wisconsin", "Wyoming", "District of Columbia")),
  vote = c("red", "red", "blue", "red", "blue", "blue", 
           "blue", "blue", "red", "blue", "blue", "red", 
           "blue", "red", "red", "red", "red", "red", 
           "blue", "blue", "blue", "blue", "blue", "red", 
           "red", "red", "red", "blue", "blue", "blue", 
           "blue", "blue", "red", "red", "red","red", "blue", 
           "blue", "blue", "red", "red","red", "red", "red", 
           "blue", "blue", "blue", "red", "blue", "red", "blue")
)

# Merge voting results with the state map
states$vote <- state_votes$vote[match(tolower(states$ID), state_votes$region)]

# Plot the map
ggplot(states) +
  geom_sf(aes(fill = vote), color = "black", size = 0.1) +
  scale_fill_manual(values = c("red" = "coral", "blue" = "skyblue"),
                    labels = c("Democrat", "Republican"),
                    name = "Party") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(
    title = "2020 U.S. Presidential Election Results by State"
  )

```


# References: